---
title: "A1 Cancers in my world"
shorttitle: "Personal data on cancer"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A1 Cancers in my world}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
  library(DT)
  data(vjc_cancer_net)
  })
 })
```

# Cancer experiences: a personal view

In 60 years of life I have been cancer-free, but many
family members and friends have had to address
the disease.

I prepared a table with some information on contacts
whose cancer experiences I know of:

```{r lktab}
datatable(vjc_cancer_net)
```

Note that the table can be sorted and searched.  You can also
alter the number of records presented.

# Questions

1. Why would a person make a table like this?
2. How could this table be made more informative?

