---
title: "A4 Using interactive tables"
shorttitle: "Interactive tables"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A4 Using interactive tables}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
library(tibble)
library(dplyr)
library(DT)
data(woncan_meta)
data(woncan)
littab = woncan |> select(MSA, `Cancer Sites`, Age.Adjusted.Rate) |> as.data.frame()
  })
 })
```

# Interactive rate tables

# A tabulation of cancer types and incidence rates across 'Metropolitan Statistical Areas'

We have formatted information from the CDC WONDER system so that
it can be searched and sorted as desired.

```{r lktab1, echo=TRUE}
datatable(littab)
```

# Exercises

A.4.1 How many areas failed to report an age-adjusted rate for prostate cancer?

A.4.2 Which state has four different areas reporting age-adjusted prostate cancer incidence exceeding 150 per 100000 men?

A.4.3 Which state has three different areas reporting female breast cancer incidence exceeding 130 per 100000 women?

A.4.4 Comment on the difficulty of making focused queries using this tabular presentation.
