---
title: "A5 Data graphics"
shorttitle: "Data graphics"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{A5 Data graphics}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
library(tibble)
library(dplyr)
library(plotly)
library(ggplot2)
library(ggbeeswarm)
library(DT)
data(woncan_meta)
data(woncan)
littab = woncan |> select(MSA, `Cancer Sites`, Age.Adjusted.Rate) |> as.data.frame()
  })
 })
```

# Visualizing variability

We have been looking at cancer incidence data in the form of tables of numbers.
We can sort, search, filter these tables fairly easily.  "Data graphics", also
known as "statistical visualization", have been devised to help us get
insights that have a different character.  One objective is to develop a
sense of variation in a quantity of interest, by aggregating data values in
various ways and presenting these aggregates in graphical form.



## Histograms
 
Histograms help visualize variability in a quantity of interest.  The range of the quantity
of interest is "binned" -- chopped up into intervals of equal width.  Midpoints of
bins are plotted on the x axis, and counts of values in bins are plotted on the y axis.

We illustrate with the CDC WONDER data on Prostate cancer incidence.

```{r lkhist, echo=TRUE}
make_hist(site="Prostate", binwidth=5, rate_table=littab)
```

We can change the visual character of the histogram display by reducing the bin width.


```{r lkhist2, echo=TRUE}
make_hist(site="Prostate", binwidth=2, rate_table=littab)
```

### Exercises

A.5.1 Based on the first histogram shown above, what is the most commonly
observed value for prostate cancer incidence (per 100,000 men) in metropolitan
statistical areas (MSAs) of the United States?  You can scroll over the display
to find values for particular MSAs.  Your answer can be a range of numbers.

A.5.2 Use the first histogram to report on the number of MSAs that report
prostate cancer incidence between 142 and 147 per 100,000 men.

A.5.3 The second histogram is produced from the same set of statistics as the
first, but uses a smaller "binwidth".  Comment on some differences between
the two displays.

A.5.4 Add a cell to your notebook using the "+" button, enter the command
`make_hist(site="Female Breast")` and press the "play" button.  What
is the most commonly reported rate of female breast cancer incidence in
MSAs in the US?

A.5.5 Hover over the leftmost block in the display.  What is the MSA with
lowest value of Female Breast cancer incidence, and what is the reported rate?

## Using a histogram to compare cancer incidence rates in two areas

By computing differences or ratios and forming the histogram, we
can assess whether there is a tendency to higher values in
one of the areas of interest.

```{r lk2, echo=TRUE}
ggplotly(make_comparison(rate_table=littab, type="ratios"))
```

More detail about individual
values can be obtained with an approach called "beeswarm" plotting.

```{r lk3, echo=TRUE}
ggplotly(make_pair(type="beeswarm", rate_table=littab ))
```

### Exercises

A.5.6 The use of ratios in the `make_comparison` display above eliminates
information about the absolute incidence rates.  Open a notebook cell and
use the commands
```
library(YESCDS)
library(plotly)
data(woncan)
ggplotly(make_comparison(rate_table=woncan, type="scatter"))
```
When you hover the mouse over points on the display, the
individual rates are presented.

True or False: the presence of a majority of points below and to the right of the
straight line implies that the incidence rates for a variety of
cancer types in Baton Rouge, LA are
more frequently higher than those in Tucson, AZ.

A.5.7 Using the values presented in the
interactive scatterplot, compute the ratio of 
incidence rates for prostate cancer in Baton Rouge
compared to those in Tucson.  Compare to the result
provided in the `make_comparison` histogram above.
