---
title: "B2 Enhancing tables with geography"
shorttitle: "Tables plus geography"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{B2 Enhancing tables with geography}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
library(tibble)
library(dplyr)
library(plotly)
library(ggplot2)
library(ggbeeswarm)
library(DT)
data(woncan_meta)
data(woncan)
library(sf)
littab = woncan |> select(MSA, `Cancer Sites`, Age.Adjusted.Rate) |> as.data.frame()
  })
 })
```

# Basic geographic data about US counties

The coordinates of central locations in US counties
are provided in `us_county_geo`.

```{r lkd3}
data(us_county_geo)
us_county_geo |> select(state, county, geometry) |> head()
```

Coordinates for metropolitan statistical areas are provided by
"statcrunch":
```{r kd4}
data(statcrunch_msa)
head(statcrunch_msa)
```

This geographic data has already been used to enhance the CDC incidence data
in `woncan`:
```{r lk5}
data(woncan)
woncan |> group_by(MSA) |> 
    summarize(lat=head(lat)[1], lng=head(lng)[1]) |> head()
```

## Exercise

B.2.1 Create a new notebook cell and run `YESCDS::table_woncan("Prostate")`

B.2.2 Use this table to find the latitude and longitude of Boston.

## Answers

B.2.1

B.2.2

# An interactive map of the US with cancer rates

After combining cancer rate data for a collection
of cancer types and metropolitan statistical areas,
we can produce a map showing variation in
cancer incidence over the United States.
Here is the example for stomach cancer:

```{r domaca}
cancer_map_usa(site="Stomach")

```

## Exercise

B.2.3 Create a new notebook cell and run `cancer_map_usa(site="Prostate", scaling=0.05)`

B.2.4 Can you identify the area with the highest incidence of prostate cancer?

## Answers

B.2.3

B.2.4

# A data graphic

Finally, to get a different view of variation in cancer rates across the United States,
run a cell with the following command.  Geographic location is lost, but the range of variation, and
details of variation can be seen very clearly in these displays.

```
browseURL("https://vjcitn.shinyapps.io/ratevariation")
```

## Exercise

B.2.5 Change the body site to "Pancreas" (remove "Brain") and find, using the histogram tab,
the areas with highest and lowest age-adjusted rates of pancreatic cancer.

## Answer

B.2.5
