---
title: "B2 Enhancing tables with geography"
shorttitle: "Tables plus geography"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{B2 Enhancing tables with geography}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
library(tibble)
library(dplyr)
library(plotly)
library(ggplot2)
library(ggbeeswarm)
library(DT)
data(woncan_meta)
data(woncan)
library(sf)
littab = woncan |> select(MSA, `Cancer Sites`, Age.Adjusted.Rate) |> as.data.frame()
  })
 })
```

# The basic geographic data about US counties

The coordinates of central locations in US counties
is provided in `us_county_geo`.

```{r lkd3}
data(us_county_geo)
us_county_geo |> select(state, county, geometry)
```

Coordinates for metropolitan statistical areas are provided by
"statcrunch":
```{r kd4}
data(statcrunch_msa)
head(statcrunch_msa)
```

This geographic data has already been used to enhance the CDC incidence data
in `woncan`:
```{r lk5}
data(woncan)
woncan |> group_by(MSA) |> 
    summarize(lat=head(lat)[1], lng=head(lng)[1]) |> head()
```

## Exercise

Use 
```
library(YESCDS)
library(DT)
data(woncan)
datatable(woncan)
```
and find the latitude and longitude of Boston.

# An interactive map of the US with cancer rates

```{r domaca}
cancer_map_usa()
```

## Exercise

Create a new notebook cell with the call to `cancer_map_usa` to produce
prostate cancer statistics.


# A data graphic

Finally, to get a different view of variation in cancer rates across the United States,
run the following cell.  Geographic location is lost, but the range of variation, and
details of variation can be seen very clearly in these displays.

```{r brow}
browseURL("https://vjcitn.shinyapps.io/ratevariation")
```

Change the body site to "Pancreas" (remove "Brain") and find, using the histogram tab,
the areas with highest and lowest age-adjusted rates of pancreatic cancer.

