---
title: "B3 Cancer in our counties"
shorttitle: "MA county cancer rates"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{B3 Cancer in our counties}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
library(tibble)
library(dplyr)
library(plotly)
library(ggplot2)
library(ggbeeswarm)
library(DT)
data(woncan_meta)
data(woncan)
library(sf)
littab = woncan |> select(MSA, `Cancer Sites`, Age.Adjusted.Rate) |> as.data.frame()
  })
 })
```

# Cancer rates in Massachusetts counties, tabulated


```{r lkd3}
datatable(MA_cancer_rate_table())
```

# Mapping the rates

```{r lkd4}
mass_cancer_map()
```

# Addressing uncertainty in the rates

The reported rates are statistics collected
over time and are adjusted using a model for
age distribution in the US.  The counts on which
the reported rates are imprecise.  Therefore
"confidence limits" are produced with the rate estimates.

Here are the prostate cancer rates in Massachusetts counties,
plotted with confidence limits.

```{r lkd5}
plotwci(MA_cancer_rate_table("prostate"), ylim=c(90,270))
```

One message from this display is that prostate
cancer in Nantucket county has an estimated age-adjusted incidence in
2014-2018 of about 200 cases per 100000 men,
but the actual rate may be lower or higher.
The uncertainty of the estimate arises from the
relatively small population of Nantucket county.

# Exercises

- What are the reported population sizes for Middlesex
and Nantucket counties?

- Run the following chunk to produce a different
representation of incidence rates for cancers
of lung and bronchus in California counties:

```
ordered_seg_cal()
```

Two statistics are presented for the 
US as a whole and for the entire state of California.

True or False: We are fairly confident that the incidence
of cancers of lung and bronchus in San Joaquin county is
lower than that of the US as a whole.

The midpoint of the plotted interval is the estimated
incidence rate.
Which California counties have incidence rates that
are around half that of the US as a whole?
