---
title: "C2 Variation in survival with cancer"
shorttitle: "Survival curves for cancers"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{C2 Variation in survival with cancer}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

```{r setup,echo=FALSE,results="hide"}
suppressMessages({
 suppressPackageStartupMessages({
  library(YESCDS)
library(tibble)
library(dplyr)
library(plotly)
library(ggplot2)
library(ggbeeswarm)
library(DT)
data(woncan_meta)
data(woncan)
littab = woncan |> select(MSA, `Cancer Sites`, Age.Adjusted.Rate) |> as.data.frame()
  })
 })
```

# The survival curve for a cohort of individuals

For a variety of factors that are mostly demanding of
further research, individuals with a given type of cancer
may exhibit longer or shorter times from diagnosis to death.

## Basic definition

When times from diagnosis to death have been recorded on
individuals in a cohort, a survival curve is a plot with
values (x, y) where x is time elapsed from diagnosis,
and y is the proportion of individuals surviving at least x units of time.

```{r doc1, echo=FALSE}
set.seed(1234)
x = rexp(200, .3/12)
library(survival)
sx = Surv(x, rep(1,200))
meddat = median(sx)
plot(survfit(sx~1), lwd=2, ylab="Survival probability", xlab="Months from diagnosis",
   axes=FALSE, xlim=c(0,168), conf.int=FALSE)
axis(2)
axis(1, at = seq(0,max(x),12L))
```

Note that at 0 months, the probability of surviving is 100%.
The probability of surviving at least 12 years (144 months)
in this cohort is very small.

## Reading off the median survival time

The median of a set of numbers is the "middle" number $m$, with
as many values smaller than $m$ as there are values larger than $m$.
Thus the median of the numbers 1, 2, 3, 4, 5 is 3.  If there is an
even number of values, we average the two middle numbers (e.g.,
for 1, 2, 3, 4, 5, 6, we would use 3.5 as the median.)

The following display shows how to read the median survival time
from a survival curve: draw the horizontal line at $y = 1/2$ and
determine the $x$ value where the intersection occurs.

```{r doc2, echo=FALSE}
plot(survfit(sx~1), lwd=2, ylab="Survival probability", xlab="Months from diagnosis",
   axes=FALSE, xlim=c(0,168), conf.int=FALSE)
axis(2)
axis(1, at = seq(0,max(x),12L))
abline(h=.5, lwd=2, lty=3)
arrows(meddat$quantile, .5, meddat$quantile, 0, lwd=2, col="blue")
```

## Reading off the five-year survival probability

```{r do5y,echo=FALSE}
plot(survfit(sx~1), lwd=2, ylab="Survival probability", xlab="Months from diagnosis",
   axes=FALSE, xlim=c(0,168), conf.int=FALSE)
axis(2)
axis(1, at = seq(0,max(x),12L))
arrows(60, .22, 0, .22, lwd=2, col="purple")
arrows(60, .0, 60, .22, lwd=2, col="purple")
```

## Exercises

For a different cohort of patients, the survival curve
is depicted below:

```{r donew,echo=FALSE}
set.seed(1234)
z = rgamma(200, 10, .15)
gg = Surv(z, rep(1,200))
plot(survfit(gg~1), conf.int=FALSE, axes=FALSE)
axis(2)
axis(1, at = seq(0,max(z),12L))
```

- True or false: For this cohort, almost all individuals live at least 1.5 years
after diagnosis

- Estimate the median survival time in months.

- What is the one-year survival probability?  What is the ten-year
survival probability?
